# Some useful tools in R

## Learning goals

-   Get to know some neat tools in R that can make coding more elegant - and easier
-   Get introduced to the `tidyverse`
-   Learn about loops
-   Get to know the `apply` function family

## An evolving language - and a lifelong learning process

The R universe is constantly evolving, offering much more now than the original base functions. Over time, modern tools and more elegant programming styles have become integral. In the upcoming chapters, some of these new tools will be introduced, along with the basics needed to use them effectively.

## The `tidyverse`

Many of the tools introduced here come from the `tidyverse` - a collection of packages developed by [Hadley Wickham](https://en.wikipedia.org/wiki/Hadley_Wickham) and his team. This collection offers many ways to improve programming skills. In this book, only the functions that are directly used will be covered. A big advantage of the tidyverse is that, with just one command - `library(tidyverse)` - all functions in the package collection become available.

## The `ggplot2` package

The `ggplot2` package, first released by [Hadley Wickham](https://en.wikipedia.org/wiki/Hadley_Wickham) in 2007, has become one of the most popular R packages because it greatly simplifies the creation of attractive graphics. The history of the package can be found [here](https://en.wikipedia.org/wiki/Ggplot2), and an introduction along with links to various tutorials is available [here](https://ggplot2.tidyverse.org/).

## The `tibble` package

A `tibble` is an enhanced version of a `data.frame` that offers several improvements. The most notable improvement is that `tibbles` avoid the common `data.frame` behavior of unexpectedly converting strings into factors. Although `tibbles` are relatively new here, they will be used throughout the rest of the book.

To create a `tibble` from a regular `data.frame` (or a similar structure), the `as_tibble` command can be used:

```{r, include = FALSE}
library(tidyverse)
```

```{r}
dat <- data.frame(a = c(1, 2, 3), b = c(4, 5, 6))
d <- as_tibble(dat)
d
```

## The `magrittr` package - pipes

`Magrittr` helps organize steps applied to the same dataset by using the pipe operator `%>%`. This operator links multiple operations on a data structure, such as a `tlibbe`, making it easier to perform tasks like calculating the sum of all numbers in the dataset:

```{r}
d %>% sum()
```

After the pipe operator `%>%`, the next function automatically takes the piped-in data as its first input, so it’s unnecessary to specify it explicitly. Additional commands can be chained by adding more pipes. This approach allows for building more complex workflows, as shown in examples later.

## The `tidyr` package

The `tidyr` package offers helpful functions for organizing data. The `KA_weather` dataset from `chillR` will be used here to illustrate some of these functions:

```{r, include = FALSE}
library(chillR)
```

```{r}
KAw <- as_tibble(KA_weather[1:10,])
KAw
```

### `pivot_longer`

The `pivot_longer` function, introduced previously, is useful for reshaping data from separate columns (like `Tmin` and `Tmax`) into individual rows. In this setup, each day’s record will have a row for `Tmin` and another for `Tmax`. This transformation is often necessary for tasks like plotting data with the `ggplot2` package. The function can be combined with a pipe for a streamlined workflow:

```{r}
KAwlong <- KAw %>% pivot_longer(cols = Tmax:Tmin)
KAwlong
```

In this example, it was necessary to specify the columns to stack. The `pivot_longer` function serves a similar purpose to the `melt` function from the `reshape2` package, which was used previously and in earlier book editions. However, `pivot_longer` is more intuitive, so it will be used throughout the remaining chapters.

### `pivot_wider`

The `pivot_wider` function allows for the opposite transformation of `pivot_longer`, converting rows back into separate columns:

```{r}
KAwwide <- KAwlong %>% pivot_wider(names_from = name) 
KAwwide
```

The `names_from` argument in `pivot_wider` specifies the column that will provide the headers for the new columns. In some cases, `pivot_wider` might work without this argument, but it’s generally recommended to include it for clarity and to ensure that the function behaves as expected, especially with more complex datasets.

### `select`

The `select` function allows users to choose a subset of columns from a `data.frame` or `tibble`:

```{r}
KAw %>% select(c(Month, Day, Tmax))
```

### `filter`

The `filter` function reduces a `data.frame` or `tibble` to just the rows that fulfill certain conditions:

```{r}
KAw %>% filter(Tmax > 10)
```

### `mutate`

The `mutate` function is essential for creating, modifying, and deleting columns in a `data.frame` or `tibble`. For example, it can be used to add new columns, such as converting `Tmin` and `Tmax` to Kelvin:

```{r}
KAw_K <- KAw %>% mutate(Tmax_K = Tmax + 273.15, Tmin_K = Tmin + 273.15)
KAw_K
```

To delete the columns created with `mutate`, you can set them to `NULL`. This effectively removes the specified columns from the `data.frame` or `tibble`:

```{r}
KAw_K %>% mutate(Tmin_K = NULL, Tmax_K = NULL)
```

Next, the original temperature values will be replaced directly with their corresponding Kelvin values. The following code will make these modifications to the specified columns:

```{r}
KAw %>% mutate(Tmin = Tmin + 273.15, Tmax = Tmax + 273.15)
```

### `arrange`

`arrange` is a function to sort data in `data.frames` or `tibbles`:

```{r}
KAw %>% arrange(Tmax, Tmin)
```

It is also possible to sort a `data.frame` or `tibble` in descending order:

```{r}
KAw %>% arrange(desc(Tmax), Tmin)
```

## Loops

In addition to the `tidyverse` functions, understanding loops is essential for efficient coding. Loops enable the repetition of operations multiple times without needing to retype or copy and paste code. They also allow for modifications to be made with each iteration. While detailed explanations on loops can be found elsewhere, the basics will be covered here.

There are two primary types of loops: **for loops** and **while loops**. For both types, it is necessary to provide instructions that determine how many times the loop will run, as well as what actions to perform during each iteration.

### *For* loops

In a for loop, explicit instructions dictate how many times the code inside the loop should be executed. This is typically done by providing a vector or list of elements, directing R to run the code for each of these elements. As a result, the number of executions corresponds to the number of elements in the vector or list. A counter is needed to track the current iteration, commonly referred to as `i`, though it can be any variable name:

```{r}
for (i in 1:3) print("Hello")
```

This command executed the code three times, producing the same output with each iteration. The structure can be made more complex by including multiple lines of code within curly brackets:

```{r}
addition <- 1

for (i in 1:3)
{
  addition <- addition + 1
  print(addition)
}
```

The code in this loop incremented an initial value of 1 by 1 in each iteration and printed the resulting value. It is important to note that R may require explicit instructions to `print` these values when the operation is embedded within a loop.

By utilizing the index `i` within the code block, additional flexibility can be introduced to the operations performed in each iteration:

```{r}
addition <- 1

for (i in 1:3)
{
  addition <- addition + i
  print(addition)
}
```

In this iteration, the respective value of `i` was added to the initial element during each run. Additionally, `i` can be utilized in more creative ways within the loop to enhance the operations being performed:

```{r}
names <- c("Paul", "Mary", "John")

for (i in 1:3)
{
  print(paste("Hello", names[i]))
}
```

The counter in a loop does not have to be numeric; it can take on various forms, including strings. By using this flexibility, the same output as generated in the previous code block can be achieved with a different formulation:

```{r}
for (i in c("Paul", "Mary", "John"))
{
  print(paste("Hello", i))
}
```

### *While* loops

A loop can also be controlled using a `while` statement, which executes the code until a specified condition is no longer met. This approach is meaningful only if the condition can change based on the operations performed within the loop:

```{r}
cond <- 5

while (cond > 0)
{
  print(cond)
  cond <- cond - 1
}
```

Once `cond` reaches 0, the starting condition is no longer satisfied, and the code will not execute again. It is important to note that while loops can lead to issues if the condition remains true regardless of the code block's execution. In such cases, the code may become stuck and will need to be manually interrupted.

## `apply` functions

In addition to loops, R offers a more efficient way to perform operations on multiple elements simultaneously. This method, which is often faster, utilizes functions from the **apply** family: `apply`, `lapply`, and `sapply`. These functions require two key arguments: the list of items to which the operation will be applied and the operation itself.

### `sapply`

When the goal is to apply an operation to a vector of elements, the simplest function to use is `sapply`. It requires two arguments: the vector (`X`) and the function to be applied (`FUN`). For illustration, a simple function called `func` will be created, which adds 1 to an input object:

```{r}
func <- function(x)
  x + 1

sapply(1:5, func)
```

The output is a vector of numbers that are each 1 greater than the corresponding elements of the input vector. When this function is applied to a list of numbers, the output becomes a matrix, although the values remain the same:

```{r}
sapply(list(1:5), func)
```

### `lapply`

To obtain a list as the output, the `lapply` function can be used. This function treats the input element `X` as a list and returns a new list containing the same number of elements as the input. Each element in the output list corresponds to the result of applying `FUN` to the respective element of the input list:

```{r}
lapply(1:5, func)
```

If the input element `X` is a list, `lapply` treats the entire list as a single input element, applying `FUN` to the whole list and returning the result as one element in the output list. An example can help clarify this behavior:

```{r}
lapply(list(1:5), func)
```

### `apply`

The basic `apply` function is designed for applying functions to arrays, allowing operations to be performed either on the rows (`MARGIN = 1`) or on the columns (`MARGIN = 2`) of the array. While this function may not be used frequently, here are some simple examples of its functionality. For further information, it is advisable to consult the help file or explore online resources, as there are many helpful materials available.

```{r}
mat <- matrix(c(1, 1, 1, 2, 2, 2, 3, 3, 3), c(3, 3))
mat
```

```{r}
apply(mat, MARGIN = 1, sum) # adding up all the data in each row
```

```{r}
apply(mat, MARGIN = 2, sum) # adding up all the data in each column
```

## `Exercises` on useful R tools

1.  Based on the `Winters_hours_gaps` dataset, use `magrittr` pipes and functions of the `tidyverse` to accomplish the following:

-   

    a)  Convert the dataset into a `tibble`

-   

    b)  Select only the top 10 rows of the dataset

```{r, include = FALSE}
library(tidyverse)
library(chillR)
library(ggplot2)
```

```{r}
WHG <- as_tibble(Winters_hours_gaps[1:10, ])
WHG
```

-   

    c)  Convert the `tibble` to a `long` format, with separate rows for `Temp_gaps` and `Temp`

To see the difference between the columns `Temp_gaps` and `Temp`, rows 279 to 302 (Julian Day 15) are used below:

```{r}
WHG <- as_tibble(Winters_hours_gaps[279:302, ])
WHGlong <- WHG %>% pivot_longer(cols = Temp_gaps:Temp)
WHGlong
```

-   

    d)  Use `ggplot2` to plot `Temp_gaps` and `Temp` as facets (point or line plot)

```{r, warning = FALSE}
ggplot(WHGlong, aes(Hour, value)) +
  geom_line(lwd = 1.5) +
  facet_grid(cols = vars(name)) +
  ylab("Temperature (°C)") +
  theme_bw(base_size = 15)
```

-   

    e)  Convert the dataset back to the `wide` format

```{r}
WHGwide <- WHGlong %>% pivot_wider(names_from = name)
WHGwide
```

-   

    f)  Select only the following columns: `Year`, `Month`, `Day` and `Temp`

```{r}
WHG %>% select(c(Year, Month, Day, Temp))
```

-   

    g)  Sort the dataset by the `Temp` column, in descending order

```{r}
WHG %>% arrange(desc(Temp))
```

2.  For the `Winter_hours_gaps` dataset, write a `for` *l*oop to convert all temperatures (`Temp` column) to degrees Fahrenheit

So that the execution of the following code does not take too long, only Julian Day 15 (rows 279 to 302) is used here. To convert the entire `Temp` column to Fahrenheit, just omit `[279:302]`

```{r}
Temp <- Winters_hours_gaps$Temp[279:302]

for (i in Temp)
{
  Fahrenheit <- i * 1.8 + 32 
  print(Fahrenheit)
}
```

3.  Execute the same operation with a function from the `apply` family

Here it is the same as in 2, just omit `[279:302]` to convert the entire `Temp` column

```{r}

x <- Winters_hours_gaps$Temp

fahrenheit <- function(x)
  x * 1.8 + 32

sapply(x[279:302], fahrenheit)
```

4.  Now use the `tidyverse` function `mutate` to achieve the same outcome

```{r}
WHG_F <- WHG %>% mutate(Temp_F = Temp * 1.8 + 32)
WHG_F
```
