# Getting temperature data

## Learning goals

-   Appreciate the need for daily temperature data
-   Know how to get a list of promising weather stations contained in an international database
-   Be able to download weather data using `chillR` functions
-   Know how to convert downloaded data into `chillR` format

## Temperature data needs

Temperature data is essential for phenology and chill modeling, as it serves as a key input for these models. Although weather data might seem easily accessible, it is often challenging to obtain. Many countries have official weather stations that record the necessary information, but access to these data is frequently restricted and expensive, despite likely being funded by taxpayers. While such costs may be manageable for small-scale studies, they quickly become prohibitive for larger analyses.

Given the urgent need to understand climate change impacts, these access limitations are a barrier to effective climate research. Such restrictions may also reduce the quality of studies that rely on comprehensive data. Ideally, high-quality, localized datasets would be available, but in their absence, some alternative databases can be used. Currently, `chillR` connects to one global and one California-specific database, with the potential to expand as more data sources become accessible.

## The Global Summary of the Day database

The [National Centers for Environmental Information (NCEI)](https://www.ncei.noaa.gov/), formerly known as the National Climatic Data Center (NCDC), offers valuable temperature data through its [Global Summary of the Day (GSOD) database](https://www.ncei.noaa.gov/access/search/data-search/global-summary-of-the-day). While accessing this data can be challenging, the bulk download section provides weather records for each station and year. However, finding specific stations can be time-consuming.

A list of weather stations is available on NOAA's website, and the process of downloading and organizing the data can be automated using the `chillR` tool. A previous user, [Adrian Fülle](https://www.gartenbauwissenschaft.uni-bonn.de/author/adrian-fulle/), developed a faster and more efficient method for this task. The `chillR` function `handle_gsod()` simplifies the process by guiding users through the necessary steps.

### `action = list_stations`

When used with this action, the `handle_gsod()` function retrieves a list of weather stations and sorts them by their proximity to specified coordinates. For example, stations near Bonn (Latitude: 50.73, Longitude: 7.10) can be identified. Additionally, a time interval, such as 1990-2020, can be set to limit the search to data from those specific years:

```{r, include = FALSE}
library(chillR)
```

```{r}
station_list <- handle_gsod(action = "list_stations",
                            location = c(long = 7.10, lat = 50.73),
                            time_interval = c(1990, 2020))
```

```{r, echo = FALSE}
library(kableExtra)
kable(station_list, align = c("c")) %>%
  kable_styling("striped", 
                position = "center", 
                font_size = 14) %>% 
  scroll_box(width = "100%",
             height = "300px",
             fixed_thead = TRUE)

```

The list provided contains the 25 closest weather stations to the specified location, ordered by their distance to the target coordinates, which is shown in the `distance` column. The `Overlap_years` column indicates the number of years for which data is available, while the `Perc_interval_covered` column shows the percentage of the target time period that is covered by the data. It's important to note that this is based solely on the start and end dates in the table. The dataset may have gaps, and these gaps can sometimes cover almost the entire record.

### `action = "download_weather"`

When used with this option, the `handle_gsod()` function downloads the weather data for a specific station, using its unique `chillR_code`, which is displayed in the respective column of the station list. Instead of manually entering the code, the function can reference the code from the station list. To download the data, the code for the 4th station in the list can be used, as it appears to cover most of the desired time period.

```{r, message = FALSE}
weather <- handle_gsod(action = "download_weather",
                       location = station_list$chillR_code[4],
                       time_interval = c(1990, 2020))
```

The result of this operation is a list with two elements. The first element (`weather[[1]]`) indicates the source of the database from which the data was retrieved. The second element (`weather[[2]]`) contains the actual weather dataset, which can be viewed here.

```{r, eval = FALSE}
weather[[1]][1:20,]
```

```{r, echo = FALSE}
library(kableExtra)
kable(weather[[1]][1:20,], align = c("c")) %>%
  kable_styling("striped", 
                position = "center", 
                font_size = 14) %>% 
   scroll_box(width = "100%",
             height = "300px",
             fixed_thead = TRUE)
```

The data appears complicated and includes unnecessary information. To simplify this, `chillR` provides a function that streamlines the record. However, this process removes several variables, including quality flags, which may indicate unreliable data. While these have been overlooked so far, there is potential for improvement.

### downloaded weather as `action` argument

This method of calling `handle_gsod()` cleans the dataset and converts it into a format that `chillR` can easily process.

```{r}
cleaned_weather <- handle_gsod(weather)
```

```{r, eval = FALSE}
cleaned_weather[[1]][1:20,]
```

```{r, echo = FALSE}
library(kableExtra)
kable(cleaned_weather[[1]][1:20,], align = c("c")) %>%
  kable_styling("striped", 
                position = "center", 
                font_size = 14) %>% 
   scroll_box(height = "300px",
             fixed_thead = TRUE)
```

The strange numbers in the dataset are due to the original database storing temperatures in degrees Fahrenheit. These values were then converted to degrees Celsius using the formula:

$Temperature[°C]=(Temperature[°F]-32)\cdot\frac{5}{9}$

This conversion process can result in awkward numbers, but it's straightforward to apply. After this conversion, the temperature records are now in a more usable format for working with `chillR`.

However, upon closer inspection, the dataset reveals significant gaps, including entire years with missing data. This issue will be addressed in the lesson on filling gaps in temperature records.

Additionally, `chillR` has a similar function for downloading data from the [California Irrigation Management Information System (CIMIS)](https://cimis.water.ca.gov/), and there is potential for more data sources to be integrated into `chillR`.

Finally, the files generated in this process will be saved for use in upcoming chapters.

```{r, eval = FALSE, warning = FALSE}
dir.create("data")
write.csv(station_list,"data/station_list.csv", row.names = FALSE)
write.csv(weather[[1]],"data/Bonn_raw_weather.csv", row.names = FALSE)
write.csv(cleaned_weather[[1]],"data/Bonn_chillR_weather.csv", row.names = FALSE)
```

## `Exercises` on getting temperature data

1.  Choose a location of interest and find the 25 closest weather stations using the `handle_gsod` function

```{r, include = FALSE}
library(chillR)
```

```{r}
station_list_Yakima <- handle_gsod(action = "list_stations",
                                   location = c(long = -120.50, lat = 46.60), 
                                   time_interval = c(1990, 2020))
```

```{r , echo = FALSE}
library(kableExtra)
kable(station_list_Yakima, align = c("c")) %>%
  kable_styling("striped", 
                position = "center", 
                font_size = 14) %>% 
  scroll_box(width = "100%",
             height = "300px",
             fixed_thead = TRUE)
```

2.  Download weather data for the most promising station on the list

```{r}
weather_Yakima <- handle_gsod(action = "download_weather",
                              location = station_list_Yakima$chillR_code[1],
                              time_interval = c(1990, 2020))
```

```{r, eval = FALSE}
weather_Yakima[[1]][1:20,]
```

```{r, echo = FALSE}
library(kableExtra)
kable(weather_Yakima[[1]][1:20,], align = c("c")) %>%
  kable_styling("striped", 
                position = "center", 
                font_size = 14) %>% 
   scroll_box(width = "100%", 
              height = "300px",
              fixed_thead = TRUE)
              
              

```

3.  Convert the weather data into `chillR` format

```{r}
cleaned_weather_Yakima <- handle_gsod(weather_Yakima) 
```

```{r, eval = FALSE}
cleaned_weather_Yakima[[1]][1:20,]
```

```{r, echo = FALSE}
library(kableExtra)
kable(cleaned_weather_Yakima[[1]][1:20,], align = c("c")) %>%
  kable_styling("striped", 
                position = "center", 
                font_size = 14) %>% 
  scroll_box(height = "300px",
             fixed_thead = TRUE)

```

```{r, warning = FALSE}
dir.create("Yakima")
write.csv(station_list_Yakima,"Yakima/station_list.csv", row.names = FALSE)
write.csv(weather_Yakima[[1]],"Yakima/raw_weather.csv", row.names = FALSE)
write.csv(cleaned_weather_Yakima[[1]],"Yakima/chillR_weather.csv", row.names = FALSE)
```
