# Historic temperature scenarios

## Learning goals

-   Understand how we can include temperature scenarios while generating synthetic weather
-   Be able to produce temperature scenarios with arbitrary change scenarios imposed
-   Understand the difference between absolute and relative temperature scenarios and the importance of baselines (for relative scenarios)
-   Learn how to produce temperature scenarios that represent past points in time from historic temperature records
-   Learn how to produce synthetic temperature scenarios for past points in time and efficiently compute agroclimatic metrics for them

## Climate change scenarios

A weather generator accessed through `chillR` can produce agroclimatic profiles for specific locations. Calibration with historical temperature data makes the generated profile representative of the climate during the calibration period. This weather generator can also simulate various climate scenarios using the `temperature_scenario` parameter in the `temperature_generation` function.

The `temperature_scenario` parameter requires a `data.frame` with columns `Tmin` and `Tmax`, each containing 12 values that specify monthly adjustments to the mean minimum and maximum temperatures. Without this parameter, the function defaults to a `data.frame` with zero adjustments, indicating no change.

A simple climate change scenario can be created by adding arbitrary values to each month’s `Tmin` and `Tmax:`

```{r}
change_scenario <- data.frame(Tmin = rep(2, 12),
                              Tmax = rep(2, 12))
```

```{r, eval = FALSE}
change_scenario
```

```{r, echo = FALSE}
library(magrittr)
library(kableExtra)
library(chillR)
kable(change_scenario, align = c("c")) %>%
  kable_styling("striped", 
                position = "center", 
                font_size = 14) 
```

```{r warning=FALSE}
Temp_2 <- temperature_generation(KA_weather,
                                 years = c(1998,2005),
                                 sim_years = c(2001,2100),
                                 temperature_scenario = change_scenario)

# As before, we'll make a data.frame that contains all
# our data, so we can take a look at it.

Temperature_scenarios <- KA_weather %>%
  filter(Year %in% 1998:2005) %>%
  cbind(Data_source = "observed") %>%
  rbind(Temp[[1]] %>% 
          select(c(Year, Month, Day, Tmin, Tmax)) %>% 
          cbind(Data_source = "simulated")
        ) %>%
  rbind(Temp_2[[1]] %>%
          select(c(Year, Month, Day, Tmin, Tmax)) %>% 
          cbind(Data_source = "Warming_2C")
        ) %>%
  mutate(Date = as.Date(ISOdate(2000, Month, Day)))
```

Due to the identical structure of this dataset to the one used in the previous lesson, the same code can be applied with only the `data.frame` name needing adjustment:

```{r message=FALSE}
ggplot(data = Temperature_scenarios, 
       aes(Date, Tmin)) +
  geom_smooth(aes(colour = factor(Year))) +
  facet_wrap(vars(Data_source)) +
  theme_bw(base_size = 20) +
  theme(legend.position = "none") +
  scale_x_date(date_labels = "%b")
```

```{r message=FALSE}
ggplot(data = Temperature_scenarios,
       aes(Date, Tmax)) +
  geom_smooth(aes(colour = factor(Year))) +
  facet_wrap(vars(Data_source)) +
  theme_bw(base_size = 20) +
  theme(legend.position = "none") +
  scale_x_date(date_labels = "%b")
```

The scenario is simplified, with future changes uniformly distributed across all months—an approach that doesn’t reflect historical patterns and is unlikely for the future. However, this method closely resembles an early attempt at modeling climate change scenarios, as illustrated in the figure from the initial publication on chilling.

![Chill scenarios for a mountain oasis in Oman according ([Luedeling et al., 2009b)](https://link.springer.com/article/10.1007/s10584-009-9581-7)(images/Luedeling_JPG_Figure_10_Future_chilling (1).jpg)

To create more realistic scenarios, specific years from historical records are used, considering both the expected typical temperature conditions of the time and actual recorded data. This approach clarifies historical climate trends by focusing on gradual climate shifts, avoiding distortions from random annual variations or extreme outliers that might obscure trend analysis.

## Making historic temperature scenarios

A long-term dataset is essential for this exercise, and the process of obtaining and preparing this data for use in `chillR` has already been covered.

```{r message=FALSE}
# download weather station list for the vicinity of Bonn
station_list <- handle_gsod(action = "list_stations",
                            location=c(7.1, 50.8))

# download weather data for Cologne/Bonn airport and convert it to chillR format
Bonn_weather <- handle_gsod(action = "download_weather",
                            location = station_list$chillR_code[1],
                            time_interval = c(1973, 2019)) %>%
  handle_gsod()

# check record for missing data
fix_weather(Bonn_weather$`KOLN BONN`)$QC

# (incidentally almost all gaps are for years covered by the KA_weather dataset)
Bonn_patched <- patch_daily_temperatures(
  weather = Bonn_weather$`KOLN BONN`,
  patch_weather = list(KA_weather))

fix_weather(Bonn_patched)$QC

# There are still 4 days missing here, out of 47 years -
# let's simply interpolate these gaps now

Bonn <- fix_weather(Bonn_patched)

Bonn_temps <- Bonn$weather
```

To create weather scenarios representing specific years like 1980, 1990, 2000, and 2010, despite using data centered around the median year of 1996 (due to gradual temperature increases from 1973 to 2019), adjustments must be made to account for historical changes. The process involves analyzing the historical temperature record to quantify shifts between the **baseline** year (the reference year of the dataset) and each target **simulation year**.

The `chillR` package provides a function, `temperature_scenario_from_records`, designed for this purpose. This function allows for the creation of temperature scenarios based on past records, facilitating the generation of simulations that accurately reflect historical conditions for specified years:

```{r}
scenario_1980 <- temperature_scenario_from_records(weather = Bonn_temps,
                                                   year = 1980)
```

This scenario involves several components:

-   `data`: A data frame containing the minimum (`Tmin`) and maximum (`Tmax`) temperatures that are representative of the year of interest.

-   `scenario_year`: The specific year for which the scenario is being created.

-   `reference_year`: This would be used if the scenario involved temperature changes relative to another year, but in this case, it is set to NA since we are working with absolute temperatures.

-   `scenario_type`: Indicates whether the scenario presents absolute temperatures for the year of interest (true in this case) or changes relative to another year.

-   `labels`: Additional information attached to the scenario, such as "running mean scenario," which provides further context or explanation about the nature of the scenario.

```{r, eval = FALSE}
scenario_1980$'1980'$data
```

```{r, echo = FALSE}
kable(scenario_1980$'1980'$data, align = c("c")) %>%
  kable_styling("striped", 
                position = "center", 
                font_size = 14)
```

The warning message received when first running the weather generator highlighted a few important points:

-   *scenario doesn’t contain named elements - consider using the following element names: ‘data’, ‘reference_year’,‘scenario_type’,‘labels’*

-   *setting scenario_type to ‘relative’*

-   *Reference year missing - can’t check if relative temperature scenario is valid*

The warning message previously encountered highlighted the missing elements in the input data frame. Scenarios require more information, and when this is absent, the temperature generation function makes assumptions about how to proceed. Specifically, the **scenario_type** was automatically set to **‘relative’**, as the function had to determine whether the scenario represented absolute or relative temperatures. Additionally, the absence of a **reference_year** meant the function couldn't identify the baseline for the relative temperature scenario.

With the updated scenario, which now includes all necessary information - such as **data**, **scenario_year**, **reference_year**, **scenario_type**, and **labels** - these warnings should no longer occur. The temperature generation function can now be run correctly with this complete set of details:

```{r}
temps_1980 <- temperature_generation(weather = Bonn_temps,
                                     years = c(1973, 2019),
                                     sim_years = c(2001, 2100),
                                     temperature_scenario = scenario_1980)
```

Despite providing all the necessary information in the updated scenario, a warning message was still encountered:

“Absolute temperature scenario specified - calibration weather record only used for simulating temperature variation, but not for the means”

The weather generator evaluated the calibration dataset by considering the variation of temperatures around the mean values. It then produced a new dataset with mean temperatures (for both **Tmin** and **Tmax**) for each month that correspond to the absolute temperature scenario provided. The variation in this generated dataset was similar to the observed temperature variation, which was the intended outcome.

To convert this absolute temperature scenario into a relative one, a baseline scenario needs to be specified for comparison. In this case, 1996 is chosen as the baseline year, since it represents the median year of the observed record. A scenario for that year can now be created as the reference for comparison:

```{r}
scenario_1996 <- temperature_scenario_from_records(weather = Bonn_temps,
                                                   year = 1996)
```

The 1996 scenario is an absolute temperature scenario. To convert it to a relative change scenario, the `temperature_scenario_baseline_adjustment` function is used, which adjusts the temperatures based on the baseline scenario.

```{r}
relative_scenario <- temperature_scenario_baseline_adjustment(
  baseline = scenario_1996,
  temperature_scenario = scenario_1980)
```

Now, a relative change scenario has been created, which contains the following elements:

-   `data`: A data frame with **Tmin** and **Tmax** columns, representing the relative changes in temperatures between 1996 and 1980 (all negative values, as 1980 was cooler than 1996).

-   `scenario_year`: 1980

-   `reference_year`: 1996

-   `scenario_type`: ‘relative’

-   `labels`: ‘running mean scenario’

The relative change scenario can now be applied similarly to previous scenarios:

```{r}
temps_1980 <- temperature_generation(weather = Bonn_temps,
                                   years = c(1973, 2019),
                                   sim_years = c(2001, 2100),
                                   temperature_scenario = relative_scenario)
```

This time, no warning message appeared because all the necessary information was provided to the `temperature_generation` function. Now, all the intended scenarios can be created easily, as most functions support vectors and lists, not just single values or data frames.

```{r}
all_past_scenarios <- temperature_scenario_from_records(
  weather = Bonn_temps,
  year = c(1980,
           1990,
           2000,
           2010))

adjusted_scenarios <- temperature_scenario_baseline_adjustment(
  baseline = scenario_1996,
  temperature_scenario = all_past_scenarios)
```

```{r, eval = FALSE}
all_past_scenario_temps <- temperature_generation(
  weather = Bonn_temps,
  years = c(1973, 2019),
  sim_years = c(2001, 2100),
  temperature_scenario = adjusted_scenarios)
```

```{r, eval=FALSE, echo=FALSE}
save_temperature_scenarios(all_past_scenario_temps, "data", "Bonn_hist_scenarios")
```

```{r, echo=FALSE}
all_past_scenario_temps <- load_temperature_scenarios("data","Bonn_hist_scenarios")
```

The `all_past_scenario_temps` object is a list containing 100 years of synthetic weather data for all the past scenarios.

To calculate chill accumulation, instead of manually applying each function to the four generated temperature scenarios, the `tempResponse_daily_list` function from `chillR` can be used, as it automates the process. Additionally, a simple frost model will be created, with the Dynamic Model and GDH Model selected for evaluation.

```{r}
frost_model <- function(x)
  step_model(x,
             data.frame(
               lower = c(-1000,0),
               upper = c(0,1000),
               weight = c(1,0)))

models <- list(Chill_Portions = Dynamic_Model,
               GDH = GDH,
               Frost_H = frost_model)
```

```{r}
chill_hist_scenario_list <- tempResponse_daily_list(all_past_scenario_temps,
                                                    latitude = 50.9,
                                                    Start_JDay = 305,
                                                    End_JDay = 59,
                                                    models = models)
```

To ensure the scenarios are available for later use in the lessons on **Making CMPI6 scenarios** and **Making CMIP5 scenarios with the ClimateWizard**, they should be saved with a clear file name that includes the place name and the start and end dates of the considered period. Before saving, any incomplete winters will be removed from the record to ensure the data is accurate and complete.

```{r}
chill_hist_scenario_list <- lapply(chill_hist_scenario_list,
                                   function(x) x %>%
                                     filter(Perc_complete == 100))

save_temperature_scenarios(chill_hist_scenario_list, "data","Bonn_hist_chill_305_59")
```

```{r message=FALSE}
scenarios <- names(chill_hist_scenario_list)[1:4]

all_scenarios <- chill_hist_scenario_list[[scenarios[1]]] %>%
  mutate(scenario = as.numeric(scenarios[1]))

for (sc in scenarios[2:4])
 all_scenarios <- all_scenarios %>%
  rbind(chill_hist_scenario_list[[sc]] %>%
          cbind(
            scenario=as.numeric(sc))
        ) %>%
  filter(Perc_complete == 100)

# Let's compute the actual 'observed' chill for comparison
actual_chill <- tempResponse_daily_list(Bonn_temps,
                                        latitude=50.9,
                                        Start_JDay = 305,
                                        End_JDay = 59,
                                        models)[[1]] %>%
  filter(Perc_complete == 100)

ggplot(data = all_scenarios,
       aes(scenario,
           Chill_Portions,
           fill = factor(scenario))) +
  geom_violin() +
  ylab("Chill accumulation (Chill Portions)") +
  xlab("Scenario year") +
  theme_bw(base_size = 15) +
  ylim(c(0,90)) +
  geom_point(data = actual_chill,
             aes(End_year,
                 Chill_Portions,
                 fill = "blue"),
             col = "blue",
             show.legend = FALSE) +
  scale_fill_discrete(name = "Scenario",
                      breaks = unique(all_scenarios$scenario)) 
```

The observed chill data should also be saved for later use in the lessons on **Making CMPI6 scenarios** and **Making CMIP5 scenarios with the ClimateWizard**. This will allow for easy access and comparison when working with those scenarios in the future.

```{r}
write.csv(actual_chill,"data/Bonn_observed_chill_305_59.csv", row.names = FALSE)
```

The chill distribution scenarios for 1980, 1990, 2000, and 2010 show minimal change, suggesting that chill deficiencies in Bonn are unlikely in the near future.

So far, all annual scenarios were based on running means of **Tmin** and **Tmax**, which is a reasonable approach given the challenges of estimating temperatures in times of accelerating climate change. However, `chillR` also offers the option of using linear regression to determine representative temperatures for a specific historical scenario. A quick comparison will reveal how this method affects the scenarios.

```{r}
temperature_means <- 
  data.frame(Year = min(Bonn_temps$Year):max(Bonn_temps$Year),
             Tmin = aggregate(Bonn_temps$Tmin,
                              FUN = "mean",
                              by = list(Bonn_temps$Year))[,2],
             Tmax=aggregate(Bonn_temps$Tmax,
                            FUN = "mean",
                            by = list(Bonn_temps$Year))[,2]) %>%
  mutate(runn_mean_Tmin = runn_mean(Tmin,15),
         runn_mean_Tmax = runn_mean(Tmax,15))


Tmin_regression <- lm(Tmin~Year,
                      temperature_means)

Tmax_regression <- lm(Tmax~Year,
                      temperature_means)

temperature_means <- temperature_means %>%
  mutate(regression_Tmin = Tmin_regression$coefficients[1]+
           Tmin_regression$coefficients[2]*temperature_means$Year,
           regression_Tmax = Tmax_regression$coefficients[1]+
           Tmax_regression$coefficients[2]*temperature_means$Year
         )

ggplot(temperature_means,
       aes(Year,
           Tmin)) + 
  geom_point() + 
  geom_line(data = temperature_means,
            aes(Year,
                runn_mean_Tmin),
            lwd = 2,
            col = "blue") + 
  geom_line(data = temperature_means,
            aes(Year,
                regression_Tmin),
            lwd = 2,
            col = "red") +
  theme_bw(base_size = 15) +
  ylab("Mean monthly minimum temperature (°C)")
```

```{r}
ggplot(temperature_means,
       aes(Year,
           Tmax)) + 
  geom_point() + 
  geom_line(data = temperature_means,
            aes(Year,
                runn_mean_Tmax),
            lwd = 2,
            col = "blue") + 
  geom_line(data = temperature_means,
            aes(Year, 
                regression_Tmax),
            lwd = 2,
            col = "red") +
  theme_bw(base_size = 15) +
  ylab("Mean monthly maximum temperature (°C)")
```

There is a noticeable difference between the running mean function (blue) and the linear regression line (red). While the running mean provides a smoother approximation, the linear regression line offers a more direct trend. This difference can be significant, and it is expected to become more pronounced as climate change continues to affect temperature patterns.

## `Exercises` on generating historic temperature scenarios

1.  For the location you chose for previous exercises, produce historic temperature scenarios representing several years of the historic record (your choice).

```{r message=FALSE}
#get a list of close-by weather stations
station_list <- handle_gsod(action = "list_stations",
                            location = c(long = -120.5, lat = 46.6),
                            time_interval = c(1973, 2023))


#download data
Yakima_weather <- handle_gsod(action = "download_weather",
                            location = station_list$chillR_code[1],
                            time_interval = c(1973, 2023)) %>%
  handle_gsod()


# check record for missing data
fix_weather(Yakima_weather$`YAKIMA AIR TERMINAL/MCALSR FIELD AP`)$QC


# filling gaps
patch_weather <-
  handle_gsod(action = "download_weather",
              location = as.character(station_list$chillR_code[c(4, 6)]),
              time_interval = c(1973, 2023)) %>%
  handle_gsod()


Yakima_patched <- patch_daily_temperatures(
  weather = Yakima_weather$`YAKIMA AIR TERMINAL/MCALSR FIELD AP`,
  patch_weather = patch_weather)


fix_weather(Yakima_patched)$QC


Yakima_temps <- Yakima_patched$weather
```

```{r}
temperature_means <- 
  data.frame(Year = min(Yakima_temps$Year):max(Yakima_temps$Year),
             Tmin = aggregate(Yakima_temps$Tmin,
                              FUN = "mean",
                              by = list(Yakima_temps$Year))[,2],
             Tmax=aggregate(Yakima_temps$Tmax,
                            FUN = "mean",
                            by = list(Yakima_temps$Year))[,2]) %>%
  mutate(runn_mean_Tmin = runn_mean(Tmin,15),
         runn_mean_Tmax = runn_mean(Tmax,15))


Tmin_regression <- lm(Tmin~Year,
                      temperature_means)

Tmax_regression <- lm(Tmax~Year,
                      temperature_means)

temperature_means <- temperature_means %>%
  mutate(regression_Tmin = Tmin_regression$coefficients[1]+
           Tmin_regression$coefficients[2]*temperature_means$Year,
         regression_Tmax = Tmax_regression$coefficients[1]+
           Tmax_regression$coefficients[2]*temperature_means$Year
  )
```

```{r}
# plot mean monthly minimum temperature of 1973 to 2023
ggplot(temperature_means,
       aes(Year,
           Tmin)) + 
  geom_point() + 
  geom_line(data = temperature_means,
            aes(Year,
                runn_mean_Tmin),
            lwd = 2,
            col = "blue") + 
  geom_line(data = temperature_means,
            aes(Year,
                regression_Tmin),
            lwd = 2,
            col = "red") +
  theme_bw(base_size = 15) +
  ylab("Mean monthly minimum temperature (°C)")
```

```{r}
# plot mean monthly maximum temperature of 1973 to 2023
ggplot(temperature_means,
       aes(Year,
           Tmax)) + 
  geom_point() + 
  geom_line(data = temperature_means,
            aes(Year,
                runn_mean_Tmax),
            lwd = 2,
            col = "blue") + 
  geom_line(data = temperature_means,
            aes(Year, 
                regression_Tmax),
            lwd = 2,
            col = "red") +
  theme_bw(base_size = 15) +
  ylab("Mean monthly maximum temperature (°C)")
```

2.  Produce chill distributions for these scenarios and plot them.

```{r}
scenario_1980 <- temperature_scenario_from_records(weather = Yakima_temps,
                                                   year = 1980)



temps_1980 <- temperature_generation(weather = Yakima_temps,
                                     years = c(1973, 2023),
                                     sim_years = c(2001, 2100),
                                     temperature_scenario = scenario_1980)

scenario_1998 <- temperature_scenario_from_records(weather = Yakima_temps,
                                                   year = 1998)

relative_scenario <- temperature_scenario_baseline_adjustment(
  baseline = scenario_1998,
  temperature_scenario = scenario_1980)

temps_1980 <- temperature_generation(weather = Yakima_temps,
                                   years = c(1973, 2023),
                                   sim_years = c(2001,2100),
                                   temperature_scenario = relative_scenario)

all_past_scenarios <- temperature_scenario_from_records(
  weather = Yakima_temps,
  year = c(1980,
           1990,
           2000,
           2010, 
           2020))

adjusted_scenarios <- temperature_scenario_baseline_adjustment(
  baseline = scenario_1998,
  temperature_scenario = all_past_scenarios)

all_past_scenario_temps <- temperature_generation(
  weather = Yakima_temps,
  years = c(1973, 2023),
  sim_years = c(2001, 2100),
  temperature_scenario = adjusted_scenarios)

save_temperature_scenarios(all_past_scenario_temps, "Yakima", "Yakima_hist_scenarios")

frost_model <- function(x)
  step_model(x,
             data.frame(
               lower = c(-1000,0),
               upper  = c(0,1000),
               weight = c(1,0)))

models <- list(Chill_Portions = Dynamic_Model,
               GDH = GDH,
               Frost_H = frost_model)

chill_hist_scenario_list <- tempResponse_daily_list(all_past_scenario_temps,
                                                    latitude = 46.6,
                                                    Start_JDay = 305,
                                                    End_JDay = 59,
                                                    models = models)

chill_hist_scenario_list <- lapply(chill_hist_scenario_list,
                                   function(x) x %>%
                                     filter(Perc_complete == 100))

save_temperature_scenarios(chill_hist_scenario_list, "Yakima","Yakima_hist_chill_305_59")

scenarios <- names(chill_hist_scenario_list)[1:5]

all_scenarios <- chill_hist_scenario_list[[scenarios[1]]] %>%
  mutate(scenario = as.numeric(scenarios[1]))

for (sc in scenarios[2:5])
  all_scenarios <- all_scenarios %>%
  rbind(chill_hist_scenario_list[[sc]] %>%
          cbind(
            scenario=as.numeric(sc))
  ) %>%
  filter(Perc_complete == 100)

# Let's compute the actual 'observed' chill for comparison
actual_chill <- tempResponse_daily_list(Yakima_temps,
                                        latitude=46.6,
                                        Start_JDay = 305,
                                        End_JDay = 59,
                                        models)[[1]] %>%
  filter(Perc_complete == 100)
```

```{r}
ggplot(data = all_scenarios,
       aes(scenario,
           Chill_Portions,
           fill = factor(scenario))) +
  geom_violin() +
  ylab("Chill accumulation (Chill Portions)") +
  xlab("Scenario year") +
  theme_bw(base_size = 15) +
  ylim(c(0,90)) +
  geom_point(data = actual_chill,
             aes(End_year,
                 Chill_Portions,
                 fill = "blue"),
             col = "blue",
             show.legend = FALSE) +
  scale_fill_discrete(name = "Scenario",
                      breaks = unique(all_scenarios$scenario)) 
```

```{r}
write.csv(actual_chill,"Yakima/Yakima_observed_chill_305_59.csv", row.names = FALSE)
```
