---
title: "Filling gaps in temperature records"
output: distill::distill_article
---

Weather data often contains gaps due to equipment malfunctions, power outages, or storage problems. These gaps create challenges for modeling agroclimatic conditions, requiring effective gap-filling methods.

## Filling Short Gaps in Daily Records

For short gaps (2-3 days), linear interpolation estimates missing values by averaging the last known and first known values around the gap. The `chillR` package provides the `interpolate_gaps()` function for this:

```{r, include=FALSE}
library(chillR)
library(tidyverse)
library(kableExtra)
library(magrittr)
```

```{r}
weather <- KA_weather %>% make_all_day_table()

Tmin_int <- interpolate_gaps(weather[,"Tmin"])
weather <- weather %>% mutate(Tmin = Tmin_int$interp, Tmin_interpolated = Tmin_int$missing)

Tmax_int <- interpolate_gaps(weather[,"Tmax"])
weather <- weather %>% mutate(Tmax = Tmax_int$interp, Tmax_interpolated = Tmax_int$missing)

KA_weather_gap <- rbind(KA_weather, c(Year = 2011,
                                      Month = 3,
                                      Day = 3,
                                      Tmax = 26,
                                      Tmin = 14)) 
```

The `fix_weather()` function can also be used to fill gaps:

```{r}
fixed_winter_days <- KA_weather_gap %>% fix_weather(start_year = 2000, 
                                                    end_year = 2011, 
                                                    start_date = 300, 
                                                    end_date = 100)
fixed_all_days <- KA_weather_gap %>% fix_weather()
```

The function returns a `weather` dataframe with interpolated data and a `QC` object summarizing interpolation quality:

```{r, eval=FALSE}
fixed_winter_days$QC
```

```{r, echo = FALSE}
kable(fixed_winter_days$QC, align = c("c")) %>%
  kable_styling("striped", 
                position = "center", 
                font_size = 14) %>% 
  scroll_box(width = "100%",
             height = "300px",
             fixed_thead = TRUE)

```

```{r, eval=FALSE}
fixed_all_days$QC
```

```{r, echo = FALSE}
kable(fixed_all_days$QC, align = c("c")) %>%
  kable_styling("striped", 
                position = "center", 
                font_size = 14) %>% 
  scroll_box(width = "100%",
             height = "300px",
             fixed_thead = TRUE)
```

A plot illustrates the effect of gap length on interpolation accuracy:

```{r}
gap_weather <- KA_weather[200:305, ]
gap_weather[ ,"Tmin_observed"] <- gap_weather$Tmin
gap_weather$Tmin[c(2, 4:5, 7:9, 11:14, 16:20, 22:27, 29:35, 
                   37:44, 46:54, 56:65, 67:77, 79:90, 92:104)] <- NA
fixed_gaps <- fix_weather(gap_weather)$weather

ggplot(data = fixed_gaps, aes(DATE, Tmin_observed)) +
  geom_line(lwd = 1.3) +
  xlab("Date") +
  ylab("Daily minimum temperature (°C)") +
  geom_line(data = fixed_gaps, aes(DATE, Tmin), col = "red", lwd = 1.3)
```

Interpolation errors increase with gap size:

```{r}
fixed_gaps[,"error"] <- abs(fixed_gaps$Tmin - fixed_gaps$Tmin_observed)

ggplot(data = fixed_gaps, aes(DATE, error)) +
  geom_line(lwd = 1.3) +
  xlab("Date") +
  ylab("Error introduced by interpolation (°C)") +
  geom_point(data = fixed_gaps[which(!fixed_gaps$no_Tmin),], aes(DATE, error), col = "red", cex = 3)
```

## Filling Long Gaps in Daily Records

For long gaps, data from nearby weather stations is used. The `patch_weather()` function in `chillR` helps with this:

```{r, echo=FALSE}
Bonn <- read.csv("data/Bonn_chillR_weather.csv")
Bonn_QC <- fix_weather(Bonn)$QC
station_list <- read.csv("data/station_list.csv")
```

```{r, eval=FALSE}
station_list <- handle_gsod(action = "list_stations",
                            location = c(7.10, 50.73),
                            time_interval = c(1990, 2020))
```

Relevant stations are downloaded:

```{r, echo=FALSE, message=FALSE}
patch_weather<-load_temperature_scenarios("data/", "patch_weather")
```

```{r, eval=FALSE}
patch_weather <- 
  handle_gsod(action = "download_weather", 
              location = as.character(station_list$chillR_code[c(2, 3, 6)]), 
              time_interval = c(1990, 2020)) %>% 
  handle_gsod()
```

Gaps are filled using `patch_daily_temperatures()`:

```{r}
patched <- patch_daily_temperatures(weather = Bonn, patch_weather = patch_weather)
```

Patch statistics are examined:

```{r, eval=FALSE}
patched$statistics[[1]]
```

```{r, echo = FALSE}
kable(patched$statistics[[1]], align = c("c")) %>%
  kable_styling("striped", 
                position = "center", 
                font_size = 14) 
```

```{r}
patched$statistics[[2]]
```

```{r, echo = FALSE}
kable(patched$statistics[[2]], align = c("c")) %>%
  kable_styling("striped", 
                position = "center", 
                font_size = 14) 
```

```{r}
patched$statistics[[3]]
```

```{r, echo = FALSE}
kable(patched$statistics[[3]], align = c("c")) %>%
  kable_styling("striped", 
                position = "center", 
                font_size = 14) 
```

To improve accuracy, mean bias and standard deviation bias limits are set:

```{r}
patched <- patch_daily_temperatures(weather = Bonn, 
                                    patch_weather = patch_weather, 
                                    max_mean_bias = 1, 
                                    max_stdev_bias = 2)
```

Final gaps are identified:

```{r}
post_patch_stats <- fix_weather(patched)$QC
```

Remaining short gaps are filled with interpolation:

```{r}
Bonn_weather <- fix_weather(patched)
```

For seasonally adjusted bias correction, `patch_daily_temps()` is used:

```{r}
patched_monthly <- patch_daily_temps(weather = Bonn, 
                                     patch_weather = patch_weather, 
                                     max_mean_bias = 1, 
                                     max_stdev_bias = 2, 
                                     time_interval = "month")
```

This function allows for interval-based bias corrections:

```{r warning=FALSE}
patched_2weeks <- patch_daily_temps(weather = Bonn, 
                                    patch_weather = patch_weather, 
                                    max_mean_bias = 1, 
                                    max_stdev_bias = 2, 
                                    time_interval = "2 weeks")
```

Using finer time intervals improves bias correction accuracy, but requires sufficient data for reliability.

## `Exercises` on filling gaps

1.  Use `chillR` functions to find out how many gaps you have in your dataset (even if you have none, please still follow all further steps)

```{r}
Yakima <- read.csv("Yakima/Yakima_chillR_weather.csv")
Yakima_QC <- fix_weather(Yakima)$QC
```

```{r, echo = FALSE}
library(kableExtra)
kable(Yakima_QC, align = c("c")) %>%
  kable_styling("striped", position = "left", font_size = 14) %>% 
  scroll_box(width = "100%",
             height = "300px",
             fixed_thead = TRUE)
```

2.  Create a list of the 25 closest weather stations using the `handle_gsod` function

```{r}
station_list_Yakima <- handle_gsod(action = "list_stations",
                                   location = c(long = -120.50, lat = 46.60),
                                   time_interval = c(1990, 2020))
```

```{r, echo = FALSE}
library(kableExtra)
kable(station_list_Yakima, align = c("c")) %>%
  kable_styling("striped", position = "left", font_size = 14) %>% 
  scroll_box(width = "100%",
             height = "300px",
             fixed_thead = TRUE)
```

3.  Identify suitable weather stations for patching gaps
4.  Download weather data for promising stations, convert them to `chillR` format and compile them in a list

```{r load_patch_weather, echo=FALSE, message=FALSE}
patch_weather<-load_temperature_scenarios("Yakima/", "patch_weather")
```

```{r, eval=FALSE}
patch_weather <-
  handle_gsod(action = "download_weather",
              location = as.character(station_list_Yakima$chillR_code[c(4, 6, 8)]),
              time_interval = c(1990, 2020)) %>%
  handle_gsod()
```

5.  Use the `patch_daily_temperatures` function to fill gaps

```{r}
patched <- patch_daily_temperatures(weather = Yakima,
                                    patch_weather = patch_weather)
```

```{r, eval = FALSE}
# Patch statistics for YRANGE OP 13 /AKIMA TRAINING CENTER
patched$statistics[[1]]
```

```{r, echo = FALSE}
kable(patched$statistics[[1]], align = c("c")) %>%
  kable_styling("striped", 
                 position = "center", 
                 font_size = 14) 
```

```{r}
# Patch statistics for HANFORD AIRPORT
patched$statistics[[2]]
```

```{r, echo = FALSE}
kable(patched$statistics[[2]], align = c("c")) %>%
  kable_styling("striped", 
                position = "center", 
                font_size = 14) 
```

```{r}
# Patch statistics for BOWERS FIELD AIRPORT
patched$statistics[[3]]
```

```{r, echo = FALSE}
kable(patched$statistics[[3]], align = c("c")) %>%
  kable_styling("striped", 
                   position = "center", 
                   font_size = 14) 
```

6.  Investigate the results - have all gaps been filled?

```{r}
write.csv(patched$weather,
          "Yakima/Yakima_weather.csv", row.names = FALSE)
```
