---
title: "Manual chill"
output: distill::distill_article
---

```{r, include = FALSE}
library(chillR)
library(kableExtra)
library(tidyverse)
```

This chapter explains how to calculate Chilling Hours using R and the `chillR` package. Chilling Hours measure the number of hours where temperatures are between 0°C and 7.2°C, which is important for certain plants to meet their cold requirements during dormancy and grow properly.

### **Data Requirements**

The calculation requires hourly temperature data, which is not always available. The `chillR` package provides tools to approximate data from daily records. In this example, the dataset *Winters_hours_gaps* is used, containing hourly temperature data recorded in 2008 from a walnut orchard in Winters, California.

### **Data Preparation**

The `chillR` package is loaded using `library(chillR)`. The relevant columns (*year, month, day, hour, temperature*) are extracted and stored in a new dataset named `hourtemps`, ensuring the correct format for calculating Chilling Hours.

```{r}
hourtemps <- Winters_hours_gaps[,c("Year",
                                   "Month",
                                   "Day",
                                   "Hour",
                                   "Temp")]
```

### **Manual Calculation of Chilling Hours**

Chilling Hours are defined as any hour where the temperature falls between 0°C and 7.2°C. In R, this is implemented using a logical condition:

```{r}
hourtemps[, "Chilling_Hour"] <- hourtemps$Temp >= 0 & hourtemps$Temp <= 7.2
```

A new column *Chilling_Hour* (TRUE/FALSE) is created to indicate whether a given hour qualifies. The total number of Chilling Hours can then be calculated using `sum(hourtemps$Chilling_Hour)`.

### **Automation with Functions**

To simplify the process, a function `CH()` was created to automatically add the *Chilling_Hour* column:

```{r}
CH <- function(hourtemps) {
  hourtemps[, "Chilling_Hour"] <- hourtemps$Temp >= 0 & hourtemps$Temp <= 7.2
  return(hourtemps)
}
```

Additionally, a function `sum_CH()` was developed to calculate the total number of Chilling Hours between two specific dates:

```{r}
sum_CH <- function(hourtemps, Start_Year, Start_Month, Start_Day, Start_Hour, 
                              End_Year, End_Month, End_Day, End_Hour) {
  hourtemps[,"Chilling_Hour"] <- hourtemps$Temp >= 0 & hourtemps$Temp <= 7.2

  Start_Index <- which(hourtemps$Year == Start_Year & hourtemps$Month == Start_Month &
                       hourtemps$Day == Start_Day & hourtemps$Hour == Start_Hour)
  End_Index <- which(hourtemps$Year == End_Year & hourtemps$Month == End_Month &
                     hourtemps$Day == End_Day & hourtemps$Hour == End_Hour)

  CHs <- sum(hourtemps$Chilling_Hour[Start_Index:End_Index])
  return(CHs)
}
```

This function uses the `which()` function to identify the relevant rows in the dataset based on the selected time range.

### **Optimization of Function Parameters**

Instead of passing year, month, day, and hour separately, compact strings in the format *YEARMODAHO* (e.g., 2008040100 for April 1, 2008, at 00:00) can be used. The function extracts values using `substr()` and converts them into numeric values.

```{r}
sum_CH <- function(hourtemps, startYEARMODAHO, endYEARMODAHO) {
  hourtemps[, "Chilling_Hour"] <- hourtemps$Temp >= 0 & hourtemps$Temp <= 7.2

  startYear <- as.numeric(substr(startYEARMODAHO, 1, 4))
  startMonth <- as.numeric(substr(startYEARMODAHO, 5, 6))
  startDay <- as.numeric(substr(startYEARMODAHO, 7, 8))
  startHour <- as.numeric(substr(startYEARMODAHO, 9, 10))

  endYear <- as.numeric(substr(endYEARMODAHO, 1, 4))
  endMonth <- as.numeric(substr(endYEARMODAHO, 5, 6))
  endDay <- as.numeric(substr(endYEARMODAHO, 7, 8))
  endHour <- as.numeric(substr(endYEARMODAHO, 9, 10))

  Start_Index <- which(hourtemps$Year == startYear & hourtemps$Month == startMonth &
                       hourtemps$Day == startDay & hourtemps$Hour == startHour)
  End_Index <- which(hourtemps$Year == endYear & hourtemps$Month == endMonth &
                     hourtemps$Day == endDay & hourtemps$Hour == endHour)

  CHs <- sum(hourtemps$Chilling_Hour[Start_Index:End_Index])
  return(CHs)
}
```

### **Application Example**

Using the function `sum_CH()`, it was calculated that between April 1st and October 11th, 2008, the walnut orchard experienced **77 Chilling Hours**:

```{r}
sum_CH(hourtemps, startYEARMODAHO = 2008040100, endYEARMODAHO = 2008101100)
```

## `Exercises` on basic chill modeling

1.  Write a basic function that calculates warm hours (\>25°C).

```{r}
WH <- function(data)
  {data[, "Warm_Hour"] <- data$Temp > 25
  return(data)
}
```

2.  Apply this function to the `Winters_hours_gaps` dataset.

```{r, eval = FALSE}
WH(Winters_hours_gaps)
```

```{r, echo = FALSE}
library(chillR)
result <- WH(Winters_hours_gaps)
kable(result[1:10, ], align = rep("c", ncol(hourtemps))) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"), 
    position = "center", 
    font_size = 14,   
    full_width = TRUE 
  )
```

3.  Extend this function, so that it can take start and end dates as inputs and sums up warm hours between these dates.

```{r}
sum_WH <- function(data, 
                   startYEARMODAHO,
                   endYEARMODAHO)
  
{data[,"Warm_Hour"] <- data$Temp > 25

startYear <- as.numeric(substr(startYEARMODAHO, 1, 4))
startMonth <- as.numeric(substr(startYEARMODAHO, 5, 6))
startDay <- as.numeric(substr(startYEARMODAHO, 7, 8))
startHour <- as.numeric(substr(startYEARMODAHO, 9, 10))

endYear <- as.numeric(substr(endYEARMODAHO, 1, 4))
endMonth <- as.numeric(substr(endYEARMODAHO, 5, 6))
endDay <- as.numeric(substr(endYEARMODAHO, 7, 8))
endHour <- as.numeric(substr(endYEARMODAHO, 9, 10))


Start_Date <- which(data$Year == startYear &
                    data$Month == startMonth &
                    data$Day == startDay &
                    data$Hour == startHour)

End_Date <- which(data$Year == endYear &
                  data$Month == endMonth &
                  data$Day == endDay &
                  data$Hour == endHour)

WHs <- sum(data$Warm_Hour[Start_Date:End_Date])
return(WHs)
}
```

Application Example:

```{r}
sum_WH(Winters_hours_gaps, startYEARMODAHO = 2008080100, 
                           endYEARMODAHO = 2008083100)
```

During the month of August 2008, from the 1st to the 31st, the walnut orchard experienced a total of 283 warm hours (defined as hours when the temperature exceeded 25°C).
