---
title: "Getting temperature data"
output: distill::distill_article
---

```{r, include=FALSE}
library(chillR)
library(tidyverse)
library(ggplot2)
library(kableExtra)
```

```{r, echo=FALSE, message=FALSE}
station_list <- read.csv("data/station_list.csv")
weather <- list(`KOLN BONN`=read.csv("data/Bonn_weather.csv"))
cleaned_weather <- list(`KOLN BONN`=read.csv("data/Bonn_chillR_weather.csv"))
```

Temperature data is crucial for phenology and chill models, as it serves as a key input for these calculations. However, accessing weather data is often challenging due to restrictions and high costs, even when the data is publicly funded.

The `chillR` package facilitates access to global and California-specific weather databases, helping researchers obtain necessary temperature records.

## Global Summary of the Day (GSOD)

The [National Centers for Environmental Information (NCEI)](https://www.ncei.noaa.gov/) provides temperature data through the [GSOD database](https://www.ncei.noaa.gov/access/search/data-search/global-summary-of-the-day). While retrieving data manually can be cumbersome, `chillR` offers a streamlined solution with the function `handle_gsod()`.

### Listing Weather Stations

To retrieve a list of weather stations near a specific location, sorted by proximity, use:

```{r, eval=FALSE}
station_list <- handle_gsod(action = "list_stations", 
                            location = c(7.10, 50.73), 
                            time_interval = c(1990, 2020))
```

This function returns a table containing station codes, available data years, and the percentage of the selected time period covered.

### Downloading Weather Data

Once a suitable station is identified, its `chillR_code` can be used to download temperature records:

```{r, eval=FALSE, message=FALSE}
weather <- handle_gsod(action = "download_weather", 
                       location = station_list$chillR_code[4], 
                       time_interval = c(1990,2020))
```

This returns a list where `weather[[1]]` contains metadata, and `weather[[2]]` holds the actual temperature dataset.

### Cleaning Weather Data

Raw GSOD data contains unnecessary variables and is recorded in Fahrenheit. `chillR` provides a function to clean and convert these records:

```{r, eval=FALSE}
cleaned_weather <- handle_gsod(weather)
```

Temperature values are converted using the formula:

$Temperature[°C]=(Temperature[°F]-32)\cdot\frac{5}{9}$

This results in a more usable dataset suitable for further analysis.

### Saving Data for Future Use

```{r, eval=FALSE}
write.csv(station_list, "data/station_list.csv", row.names=FALSE)
write.csv(weather[[1]], "data/Bonn_raw_weather.csv", row.names=FALSE)
write.csv(cleaned_weather[[1]], "data/Bonn_chillR_weather.csv", row.names=FALSE)
```

## `Exercises` on getting temperature data

```{r, echo=FALSE, message=FALSE}
station_list_Yakima <- read.csv("Yakima/station_list.csv") 
weather_Yakima <- list(`YAKIMA AIR TERMINAL/MCALSR FIELD AP`=read.csv("Yakima/Yakima_weather.csv")) 
cleaned_weather_Yakima <- list(`YAKIMA AIR TERMINAL/MCALSR FIELD AP`=read.csv("Yakima/Yakima_chillR_weather.csv"))
```

1)  Choose a location of interest and find the 25 closest weather stations using the `handle_gsod` function

```{r}
station_list_Yakima <- handle_gsod(action = "list_stations",
                                   location = c(long = -120.50, lat = 46.60), 
                                   time_interval = c(1990, 2020))
```

```{r , echo = FALSE}
library(kableExtra)
kable(station_list_Yakima, align = c("c")) %>%
  kable_styling("striped", 
                position = "center", 
                font_size = 14) %>% 
  scroll_box(width = "100%",
             height = "300px",
             fixed_thead = TRUE)
```

2)  Download weather data for the most promising station on the list

```{r, eval=FALSE}
weather_Yakima <- handle_gsod(action = "download_weather",
                              location = station_list_Yakima$chillR_code[1],
                              time_interval = c(1990, 2020))
```

```{r, eval = FALSE}
weather_Yakima[[1]][1:20,]
```

```{r, echo = FALSE}
library(kableExtra)
kable(weather_Yakima[[1]][1:20,], align = c("c")) %>%
  kable_styling("striped", 
                position = "center", 
                font_size = 14) %>% 
   scroll_box(width = "100%", 
              height = "300px",
              fixed_thead = TRUE)
              
              

```

3)  Convert the weather data into `chillR` format

```{r, eval=FALSE}
cleaned_weather_Yakima <- handle_gsod(weather_Yakima) 
```

```{r, eval = FALSE}
cleaned_weather_Yakima[[1]][1:20,]
```

```{r, echo = FALSE}
library(kableExtra)
kable(cleaned_weather_Yakima[[1]][1:20,], align = c("c")) %>%
  kable_styling("striped", 
                position = "center", 
                font_size = 14) %>% 
  scroll_box(height = "300px",
             fixed_thead = TRUE)

```

```{r, warning = FALSE}
dir.create("Yakima")
write.csv(station_list_Yakima,"Yakima/station_list.csv", row.names = FALSE)
write.csv(weather_Yakima[[1]],"Yakima/raw_weather.csv", row.names = FALSE)
write.csv(cleaned_weather_Yakima[[1]],"Yakima/chillR_weather.csv", row.names = FALSE)
```
